Efficient forwards-time simulation with tree sequences

### 2017-11-12 forwards_paper.tex
moar edits

Suppose we have a tree sequence consisting of a set of nodes and edges as
described earlier, and we have some subset of the nodes in this tree sequence
that we are interested in (our `samples').  We wish to reduce the input tree sequence
into the minimum representation of the topologies that include the specified
samples. The output tree sequence must have the following properties:
\begin{enumerate}
\item we must have the same marginal trees with respect to the samples as
the input;
\item within the marginal trees, all non-sample vertices must has at least
two children (i.e., unary tree vertices are removed);
\item any nodes and edges not reachable from the sampled nodes are removed;
\item there are no adjacent redundant edges, i.e, pairs of edges $(\ell, x, p,
c)$ and $(x, r, p, c)$ which can be represented with a single edge
$(\ell, r, p, c)$.
\end{enumerate}

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        A simple example of the method.
        \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
        It is recorded, in forwards time,
        in 10 node records (one for each individual)
        and 14 edge records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations in each meiosis.
        The individuals $B$, $C$, and $D$ inherit clonally from $A$.
        \textbf{Center:} the five distinct trees relating all individuals to each other
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red.
        \textbf{Bottom:} tables recording the tree sequence after simplification
        with nodes $I$ and $J$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red,
        which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

In the current context of running forwards in time simulations, the tables of
nodes and edges that we record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of the current day population. We can use the process of simplification here to
reduce this down to the minimal set of nodes and edges required to represent
the information that we are interested in. We can also use simplification if  we have some very
large tree sequence representing a large dataset and we wish to extract the
information relevant to a subset of the samples.

Roughly, simplification works by tracing ancestry from the samples backwards
through the recorded history, adding node and edge records to the output
only when coalescent events are reached. This works exactly as in \msprime,
allowing substantial re-use of algorithms; the main difference being that
parental choice, mutations, and recombination locations are determined by the
input tree sequence rather than randomly generated. An example is shown in
Figure \ref{fig:method_diagram}.

\begin{figure}
    \begin{center}
        \includegraphics{simplify-state-diagram}
    \end{center}
    \caption{
        A depiction of the state of the simplification algorithm.
        \label{fig:simplify_state}
    }
\end{figure}

The approach that we take is based on Hudson's algorithm for simulating
the coalescent with
recombination~\citep{hudson1983properties,kelleher2016efficient}. The main
state of the algorithm is a set of ancestral segments; each segment $(\ell, r,
u)$ represents a tract of ancestral material in which the node $u$ is
associated with the genomic interval $[\ell, r)$. In the same manner
as Hudson's algorithm, we represent the set of extant lineages as a
set of individuals, where each ancestral individual is composed of a
linked list of these segments. The node $u$ in each ancestral sequence
is the node ID associated with this segment in the output tree sequence.
We maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
the input tree sequence.

@@

\subsection*{Data structures:}
\jk{Moved from above.}

>>>>

Suppose that we are only interested in a subset of the nodes (our `samples')
of a tree sequence,
and so wish to reduce this input tree sequence
into the minimum representation of the topologies that include the specified
samples. The output tree sequence must have the following properties:
\begin{enumerate}
\item All marginal trees must match the subtree induced by the samples of the corresponding tree in the input tree sequence.
\item Within the marginal trees, all non-sample vertices must has at least
two children (i.e., unary tree vertices are removed), unless the vertex is a sample.
\item Any nodes and edges not ancestral to the sampled nodes are removed.
\item There are no adjacent redundant edges, i.e, pairs of edges $(\ell, x, p,
c)$ and $(x, r, p, c)$ which can be represented with a single edge
$(\ell, r, p, c)$.
\end{enumerate}

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        A simple example of the method.
        \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
        It has 10 node records (one for each individual)
        and 14 edge records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations in each meiosis --
        for instance, $C$ and $D$ were parents to $F$,
        who inherited the left 70\% of the chromosome from $D$ and the remainder from $C$.
        The individuals $B$, $C$, and $D$ inherit clonally from $A$.
        \textbf{Center:} the five distinct trees relating all individuals to each other
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red.
        \textbf{Bottom:} tables recording the tree sequence after simplification
        with nodes $I$ and $J$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red.
        % which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

The tree sequences produced by forwards simulations, 
record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of the current population.
Simplification is essential to reduce this 
to a manageable quantity that still contains all
the information that we are interested in. 
Simplification is also useful if we have a
large tree sequence representing a large dataset and we wish to extract the
information relevant to a subset of the samples.

The approach that we take is based on Hudson's algorithm for simulating
the coalescent with recombination~\citep{hudson1983properties,kelleher2016efficient};
the implementaiton of simplify parallels the implementation of Hudson's algorithm in \msprime.
Conceptually, this works by 
(a) beginning by painting the chromosome each sample a distinct color;
(b) moving back through history, 
copying the colors of each chromosome to the portions of its' parental chromosomes
from which it was inherited;
(c) each time we would paint two colors in the same spot (a coalescence),
record that information as an edge and instead paint a brand-new color; 
and
(d) once all colors have coalesced on a given segment,
stop propagating it.
This ``paint pot'' description misses some details --
for instance, we must ensure that all coalescing segments in a given individual
are assigned the \emph{same} new color --
but is reasonably close.
The process is depicted in Figure~\ref{fig:simplify_state}.

\begin{figure}
    \begin{center}
        \includegraphics{simplify-state-diagram}
    \end{center}
    \caption{
        A depiction of the state of the simplification algorithm
        at each point in time,
        in the example of Figure~\ref{fig:method_diagram}.
        \label{fig:simplify_state}
    }
\end{figure}

More concretely,
the algorithm works by moving back through time,
processing each edge in the input tree sequence in chronological order.
The main state of the algorithm at each point in time is a set of ancestral segments:
each segment $(\ell, r, u)$ represents a tract of ancestral material in which the node $u$ is
associated with the genomic interval $[\ell, r)$. 
These segments are stored in a collection of linked lists,
one list of segments for each ancestral lineage present at that time.
(These lineages are the ``colors'' above.)
The node $u$ in each ancestral sequence
is the node ID associated with this segment in the output tree sequence.
We maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
the input tree sequence.

@@

%%%%%%%%%%%%%%
\subsection*{Simplification}

Here we describe the simplification algorithm.
\plr{I am trying it out to see what it looks like.}

\begin{taocpalg}{S}{Simplify a tree sequence}
{Simplifies the input tree sequence.
}

\algstep{S0.}{Initialize.}{
    Set the samples.
}

\algstep{S1.}{Input parent loop.}{Loop over parents, chronologically.
}

\algstep{S2.}{Process parent edges.}{Loop over edges associated with this parent, removing ancestry from each child.
}

\algstep{S3.}{Merge ancestors.}{Merge the resulting segments into a new ancestor for this parent.
}

This uses
\begin{taocpalg}{R}{Remove ancestry}
    {Remove ancestry on a segment from an ancestor.}

\end{taocpalg}
and
\begin{taocpalg}{M}{Merge ancestors}
    {Merge a collection of segments and create a new ancestor.}

%%%%%%%%%%%%%%
\subsection*{Data structures:}
\jk{Moved from above.}
